
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```





###Public Health and Economoic Consequences of Severe Weather in the U.S. 

</br>


###Synopsis
This analysis aims to answer two key questions about severe weather and related
atmospheric and meteorlogical activity in the U.S.:

- Which types of events are most harmful with respect to population health?
- Which types of events have the greatest economic consequences?

The  focus will be on determining the top 10 events separately for both fatalaties 
and injuries. Likewise, to determine the top 10 events with economic consequences
we will analyze property and crop damage separetly.

The decision was made to take property and crop damage estimates as-is and not 
adjust amounts for inflation. Further analysis that adjusts for inflation and/or 
analyzes data by year may yield different results.

</br>


###Data Repository
Data was obtained from the U.S. National Oceanic and Atmospheric Administration's 
(NOAA) storm database between 1950 and November 2011. Data may be downloaded here:
 <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2>.

</br>


###Data Definitions and Documentation
- National Weather Service Storm Data Documentation may be found here: <https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf>
- National Climatic Data Center Storm Events FAQ may be found here:
<https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf>

With this analysis limited to health and economic damages across the U.S. as a whole
and without regard to the timing or year of the even, we need only consider the 
columns below from the dataset:
- EVTYPE: type of event (i.e. Hail, Tornado, Flash Flood)
- FATALITIES: deaths related to the event
- INJURIES: injuries related to the event
- PROPDMG: amount of property damage
- PROPDMGEXP: multiplier for property damage. (0-1: 1; H: 100; K: 1,000; M: 1,000,000; B: 1,000,000,000)
- CROPDMG: Base amount of crop damage.
- CROPDMGEXP: Base multiplier for crop damage. (0-1: 1; H: 100; K: 1,000; M: 1,000,000; B: 1,000,000,000)

</br>


###Data Processing
```{r}
# load required libraries
library(knitr)
library(ggplot2)

#change this line to your workind directory if reproducing this research
setwd("/Users/schre/DataScience/wd/repro_research/Project2")

#read in data
if(!file.exists("./data")){dir.create("./data")}
if (!file.exists('data/repdata%2Fdata%2FStormData.csv.bz2')) {
      fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
      download.file(fileurl, 'data/repdata%2Fdata%2FStormData.csv.bz2')
}
storm.data <- read.csv("data/repdata%2Fdata%2FStormData.csv.bz2", strip.white = TRUE,
                 stringsAsFactors = FALSE)
```

First, we will eliminate any columns not required:
```{r}
#limit data to only those columns we need
storm.data <- subset(storm.data, select = c(EVTYPE,FATALITIES,INJURIES,PROPDMG,
                  PROPDMGEXP,CROPDMG,CROPDMGEXP))

```

Second, to make better sense of the property and crop damage numbers, we will take
the cost x the multiplier (H, K, M, B) and create a new column called "propdmg"
and "cropdmg":
```{r}
storm.data$propdmg <- 0
storm.data[storm.data$PROPDMGEXP == "H", ]$propdmg <- 
  storm.data[storm.data$PROPDMGEXP == "H", ]$PROPDMG * 10^2

storm.data[storm.data$PROPDMGEXP == "K", ]$propdmg <- 
  storm.data[storm.data$PROPDMGEXP == "K", ]$PROPDMG * 10^3

storm.data[storm.data$PROPDMGEXP == "M", ]$propdmg <- 
  storm.data[storm.data$PROPDMGEXP == "M", ]$PROPDMG * 10^6

storm.data[storm.data$PROPDMGEXP == "B", ]$propdmg <- 
  storm.data[storm.data$PROPDMGEXP == "B", ]$PROPDMG * 10^9

storm.data$cropdmg <- 0
storm.data[storm.data$CROPDMGEXP == "H", ]$cropdmg <- 
  storm.data[storm.data$CROPDMGEXP == "H", ]$CROPDMG * 10^2

storm.data[storm.data$CROPDMGEXP == "K", ]$cropdmg <- 
  storm.data[storm.data$CROPDMGEXP == "K", ]$CROPDMG * 10^3

storm.data[storm.data$CROPDMGEXP == "M", ]$cropdmg <- 
  storm.data[storm.data$CROPDMGEXP == "M", ]$CROPDMG * 10^6

storm.data[storm.data$CROPDMGEXP == "B", ]$cropdmg <- 
  storm.data[storm.data$CROPDMGEXP == "B", ]$CROPDMG * 10^9
```
 
 
 
###Results
Between 1950 to November 2011 the U.S. has seen a total of 15,145 fatalaties and 
140,528 injuries:
```{r}
sum(storm.data$FATALITIES)
sum(storm.data$INJURIES)
```

Furthermore, property and crop damage is an estimated $427 and $49 billion respectively:
```{r}
paste("$",format(sum(storm.data$propdmg), big.mark=","),sep="")
paste("$",format(sum(storm.data$cropdmg), big.mark=","),sep="")
```



In order to break down these numbers we first must recognize that there are event
types are not standardized. The dataset constains 985 distinct event types:
```{r}
unique(storm.data$EVTYPE)
```

Many of these are similar, vague terms or even combinations of some event types. 
For example, "Thunderstorm", "Wind" and "Flood" are common events, but we can find both
"Flash Flooding/Thunderstorm", "Thunderstorm Wind", etc. In fact, there are 101 
event types containing the word 'flood':

```{r}
unique(grep("flood", tolower(storm.data$EVTYPE), value=T))
```


Our strategy will be to view the top 20 events and using regular expressions we'll
combine similar event types. An initial plot of the top 20 events then looks like so: 
```{r}
#First we will find the top 20 event types in order to derive the EVTYPE names we
#will later use in our regular expressions

storm.data.injuries <- aggregate(INJURIES ~ EVTYPE, data=storm.data, sum)
storm.data.injuries <- 
      storm.data.injuries[order(-storm.data.injuries$INJURIES), ][1:20, ]

ggplot(storm.data.injuries, aes(x = EVTYPE, y = INJURIES)) + 
  geom_bar(stat = "identity", fill = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("EVENT TYPE") + ylab("INJURIES") +
  ggtitle("Top 20 Weather Events by Number of Injuries")

```

From this we see that we can combine the following events:
- FLASH FLOOD and FLOOD
- THUNDERSTORM WIND, THUNDERSTORM WINDS, and TSTM WIND
- EXCESSIVE HEAT and HEAT
- BLIZZARD, HEAVY SNOW, ICE STORM, and WINTER STORM


```{r}
#Standardize event types
storm.data$EVTYPE[grep("THUN|TSTM", storm.data$EVTYPE, ignore.case = TRUE)] <- "THUNDERSTORM"
storm.data$EVTYPE[grep("FLOOD", storm.data$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
storm.data$EVTYPE[grep("TORN", storm.data$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
storm.data$EVTYPE[grep("BLIZZARD|SNOW|ICE|FREEZE", storm.data$EVTYPE, ignore.case = 
                             TRUE)] <- "WINTER STORM"
storm.data$EVTYPE[grep("HURRICANE", storm.data$EVTYPE, ignore.case = TRUE)] <- "HURRICANE"
storm.data$EVTYPE[grep("HEAT", storm.data$EVTYPE, ignore.case = TRUE)] <- "HEAT"
```


Next we will re-plot using the top 10 events for impacts on health. In this analysis 
injuries and fatalities will simply be added together, no distinction will be made
to their relative effects, as the relative serious of injuries alone cannot be determined.

```{r}
storm.data.health <- setNames(aggregate(INJURIES+FATALITIES ~ EVTYPE, 
      data=storm.data, sum), c("EVTYPE", "VALUE"))

storm.data.health <- 
      storm.data.health[order(-storm.data.health$VALUE), ][1:10, ]
storm.data.health <- transform(storm.data.health, EVTYPE = reorder(EVTYPE, VALUE))

ggplot(storm.data.health, aes(x = EVTYPE, y = VALUE)) + 
    geom_bar(stat = "identity", fill = "red") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("1950 - late 2011") + ylab("INJURIES + FATALATIES") +
    ggtitle("Top 10 Weather Events Impacting Public Health")

```


Finally, we will use a similar methodology to plot the top 10 events for impacts 
the economy, only this time we will chart property damage


```{r}
storm.data.econ <- setNames(aggregate(propdmg+cropdmg ~ EVTYPE, 
      data=storm.data, sum), c("EVTYPE", "VALUE"))

storm.data.econ <- 
      storm.data.econ[order(-storm.data.econ$VALUE), ][1:10, ]
storm.data.econ <- transform(storm.data.econ, EVTYPE = reorder(EVTYPE, VALUE))

ggplot(storm.data.econ, aes(x = EVTYPE, y = VALUE/10^9)) + 
    geom_bar(stat = "identity", fill = "red") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("1950 - late 2011") + ylab("PROPERTY + CROP DAMAGE ($ in billions)") +
    ggtitle("Top 10 Weather Events Impacting the U.S. Economy")

```